\chapter{Introduction}
Persistence is the topic dealing with the general availability of previous
versions of a data structure for various purposes. The antonym to ``persistent''
is in this context called ``ephemeral'', which means that no previous version of
a data structure is available. In the 1989 article by Driscoll et al.
\cite{Driscoll198986}, it is described how a data structure may be made
persistent% in two different ways: partially or fully
.

With partially persistent data structures, any previous version is accessible,
but not modifiable, and as such changes to the data structure can only be made
on the most recent version.

% With fully persistent data structures, it is possible to make changes to a
% previous version of the data structure, thus creating an alternative branch in
% the `history' of the data structure.

Partially persistent data structures can be compared with ``rollback'' databases
as defined in \cite{10.1109/AFIPS.1987.11}. ``Rollback'' databases allow reading
from any previous version of the data structure, but only allow changing the
most recent version. Driscoll et al. describe a method for making any bounded
in-degree data structure partially persistent with $O(1)$ amortized space
overhead per version and $O(1)$ amortized overhead for access of any version and
operations on the most recent version. This method is called Node Copying.

My thesis will analyse and compare Node Copying and Rollback as two different
approaches to obtaining partial persistence, using a doubly linked list as an
example data structure.

% \section{Chapter overview}
% In the \nameref{chapter:method} chapter, the theory behind the two approaches is
% analysed, and possible ways to of optimizing the performance of the Rollback
% approach are investigated. Time and space complexities are shown for retrieving
% a desired version and for adding a new version by changing the most recent one.
% 
% In the \nameref{chapter:empirical-analysis} chapter, experiments run with
% implementations of the approaches are described, and findings based on the
% resulting data are discussed and explained. The experiments cover two different
% usage scenarios. The purpose of the empirical analysis is to form a practical
% foundation based on which one or the other approach may be recommended when
% partial persistence is needed in practice.
% 
% In the \nameref{chapter:conclusion} chapter, overall conclusions and
% recommendations are made based on the theoretical and empirical analyses, and
% suggestions are given for future work.

\section{Scope}

In my thesis, only data structures with bounded in-degree will be considered,
and the scope is limited to partially persistent data structures.

The thesis is mainly concerned with making a doubly linked list partially
persistent.

When discussing time and space complexity, the RAM model will be used, and
big-O notation is used when deriving and comparing such complexities.
